Measuring Interaction Design

### 2015-09-24 paper.tex
first version of background section completed.

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\newcommand{\mytodo}[1]{\todo[inline]{#1}}
\usepackage{soul}
\usepackage{paralist}
\usepackage{listings}

@@

Even worse, there might be questions that cannot be cannot be easily
answered. Given one or more potential designs and some usage
scenarios, interesting questions could include: ``How many different
routes can be followed by the user to carry out the scenario?'',
``Which are the shortest ones?'', ``If a user makes a mistake, would
he or she be able to recover?'', ``How many steps would the recovery
require?''. In a different setting, when designing and evaluating
embedded UIs (such as when dealing with plane's
cockpits~\cite{conversy07}), other relevant questions might include
``How would the above properties change if we add a certain a
widget?'', or ``... if we replace a widget with another?''.  At the
moment, these straightforward questions are quite complicated to
answer. In fact, they require development of prototypes, inspecting
them, manual tracking of which screens and widgets are used at which
stage, and a systematic manual analysis.

@@

Our contribution consists of
\begin{inparaenum}[(a)]
\item the idea of using a model to specify interaction scenarios,
\item the development of a tool (MIG-TOOL, Measuring Interaction
  Graphs Tool) that transforms models and scenario specifications into
  interaction paths, and
\item the definition of metrics that provide concise, precise and
  objective measures of a design.
\end{inparaenum}
The examples included in the paper show that among four web mail
applications, and with respect to a typical usage scenario, Gmail is
the most efficient and flexible UI, with the best ability to recover
from user errors, but only for users that possess a certain level of
proficiency. In fact, Gmail has the largest number of shortest paths
(when users are supposed to make 0 or 1 error); when users make 2 or
more errors the number of paths drops significantly, reducing the
error-proneness of the UI; on the best case, Gmail features also the
shortest paths, requiring 13\% fewer steps; however, the probability
that a user hits an optimal path with Gmail is 10 times smaller than
the best of the other applications, and the probability that a random
walker hits a state that is not due to an error is 10\% smaller than
the best of the other applications.
%
These values suggest that Gmail offers more ways to accomplish tasks
included in the scenario, that comparably more of these ways do not
involve extra steps, that they require fewer steps, and that it might
be more difficult for novice users to exploit the most efficient
ways. Further inspections show that some differences are due to the
slightly different interaction structures adopted for uploading
messages. If Gmail didn't exist yet, by using MIGTOOL its designers
could obtain these answers well before developing prototypes and
performing usability studies.

@@

(right now these are random notes)

SwingState: work done by 
Caroline Appert\cite{appert2008swingstates}

in java, FSA used as a conceptual framework to write the code of
widgets so that events and event handlers in the ui can more easily be
conceived, developed and verified.

They say: StateCharts were used in the StateMaster User Interface
Management System [40], and a variant of them specifically tailored to
designing user interfaces was used in the more recent HsmTk toolkit
[18]. StateCharts however are significantly more complicated and hard
to learn than plain state machines, and our experience is that user
interface designers and developers have difficulties exploiting their
power. Other approaches include Petri Nets [41], which have also been
used to specify user interfaces, for example in the PetShop system
[42]. Here too, the learning curve is steep, making the adoption of
such a model by developers difficult.

With SwingStates, any number of state machines can run simulateneously. A state machine
can be active, i.e., handling the events it receives, or inactive,
i.e., ignoring events

We adopt a similar apporach, but using UML state machine (which are
more powerful than FSA) for conceiving, guiding development, analysis
and verification of the behavior of the entire app.

Nebeling says:\cite{nebeling2011metrics}

Despite the fact that screen sizes and average screen resolutions have
dramatically increased over the past few years, little attention has
been paid to the design of web sites for large, high-resolution
displays that are now becoming increasingly used both in enterprise
and consumer spaces. We present a study of how the visual area of the
browser window is currently utilised by news web sites at different
widescreen resolutions. The analysis includes measurements of space
taken up by the article content, embedded ads and the remaining
components as they appear in the viewport of the web browser. The
results show that the spatial distribution of page elements does not
scale well with larger viewing sizes, which leads to an increasing
amount of unused screen real estate and unnecessary scrolling. We
derive a number of device-sensitive metrics to measure the quality of
web page layout in different viewing contexts, which can guide the
design of flexible layout templates that scale effectively on large
screens.

Stephane Conversy\cite{conversy07} says: (friend of nicolas roussel in
the committee)

ARINC 661 provides precise information for communication protocol
between application (called User Applications) and user interface
components (called widgets) as well as precise information about the
widgets themselves. However, in ARINC 661, no information is given
about the behaviour of these widgets and about the behaviour of an
application made up of a set of such widgets.

The purpose of ARINC 661 specification (ARINC 661, 2002) is to define
interfaces to a Cockpit Display System (CDS) used in interactive
cockpits that are now under

application of a formal description technique to the various elements
of ARINC 661 specification within an industrial project. This formal
description technique called Interactive Cooperative Objects defines
in a precise and non-ambiguous way all the elements of ARINC 661
specification. The application of the formal description techniques is
shown on an interactive application called MPIA (Multi Purpose
Interactive Application). Within this application, we present how ICO
are used for describing interactive widgets, User Applications and
User Interface servers (in charge of interaction techniques). The
emphasis is put on the model-based management of the feel of the
applications allowing rapid prototyping of the external presentation
and the interaction techniques.

The Interactive Cooperative Objects (ICOs) formalism is a formal
description technique dedicated to the specification of interactive
systems [4, 11]. It uses concepts borrowed from the object-oriented
approach to describe the structural or static aspects of systems, and
uses high-level Petri nets [8] to describe their dynamic or
behavioural aspects.

The Interactive Cooperative Objects (ICOs) formalism is a formal
description technique dedicated to the specification of interactive
systems [4, 11]. It uses concepts borrowed from the object-oriented
approach to describe the structural or static aspects of systems, and
uses high-level Petri nets [8] to describe their dynamic or
behavioural aspects.

Voida\cite{voida} says:
We describe an alternative model for organizing the
desktop interface-activity-based computing-and
identify a series of high-level system requirements for
interfaces that use activity as their primary organizing
principle.

We could mention this work when we introduce scenarios, and say that
people inventing these new metaphors could take advantage of our
metrics if they model 2+ apps.

Eg. they say: Requirement 2. Activity-based systems should provide
lightweight mechanisms to create, change, and alter
activities, since heavyweight interaction techniques are
likely to deter adoption and use.

eg Requirement 7. Because information sharing is a
“common case” in knowledge work, lightweight sharing
capabilities should be integrated directly as a first-class
interaction technique.

@@

Notice that a scenario specifies only the desired occurrences of user
actions, not the necessary ones. For example, the model specifies that
in order to view a conversation while reading another one, one has to
\texttt{goBack} to the inbox first. It is the task of MIGTOOL to
unfold cyles in the graph and compute all the  paths that connect the
desired user actions.

@@

feasibility:

sensitivity with respect to models

dire di horrocks \cite{horrocks99}

dire di agile modeling \cite{ambler02}

\hl{dire} di SEPARATING concerns, ad es M-V-C, per vedere le
  implicazioni di ciascuna e non soffrire di ``saturation'', vedi adams

valdiity of results:

dire che le ns analisi sono prive di assunzioni su conoscenze
abilita'; le probability sono puramente sulla base delle frequenze.

dire di cogtools: noi non prevediamo i tempi (richiede modeling
psicologico e dei specifici widgets). pero' forniamo un'analisi che
puo' guidare chi applica cogtool.

MIGTOOL is part of a suite of model-based tools for designing,
analysing and testing user interfaces developed by the company
ANONYMIZED. Development of a web-based user interface of MIGTOOL is
underway; the company plans to make it freely available for research
purposes.

>>>>

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\newcommand{\mytodo}[1]{\todo[inline]{#1}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\migtool}{MIG-Tool\ }
\usepackage{soul}
\usepackage{paralist}
\usepackage{listings}

@@

Even worse, there might be questions that cannot be easily
answered. Given one or more potential designs and some usage
scenarios, interesting questions could include: ``How many different
routes can be followed by the user to carry out the scenario?'',
``Which are the shortest ones?'', ``If a user makes a mistake, would
he or she be able to recover?'', ``How many steps would the recovery
require?''. In a different setting, when designing and evaluating
embedded UIs (such as when dealing with plane's
cockpits~\cite{conversy07}), other relevant questions might include
``How would the above properties change if we add a certain a
widget?'', or ``... if we replace a widget with another?''.  At the
moment, these straightforward questions are quite complicated to
answer. In fact, they require development of prototypes, inspecting
them, manual tracking of which screens and widgets are used at which
stage, and a systematic manual analysis.

@@

Our contribution consists of
\begin{inparaenum}[(a)]
\item the idea of using a model to specify interaction scenarios,
\item the development of a tool (MIG-TOOL, Measuring Interaction
  Graphs Tool) that transforms models and scenario specifications into
  interaction paths, and
\item the definition of metrics that provide concise, precise and
  objective measures of a design.
\end{inparaenum}
The examples included in the paper show that among four web mail
applications, and with respect to a typical usage scenario, Gmail is
the most efficient and flexible UI, with the best ability to recover
from user errors, but only for users that possess a certain level of
proficiency. In fact, Gmail has the largest number of shortest paths
(when users are supposed to make 0 or 1 error); when users make 2 or
more errors the number of paths drops significantly, reducing the
error-proneness of the UI; on the best case, Gmail features also the
shortest paths, requiring 13\% fewer steps; however, the probability
that a user hits an optimal path with Gmail is 10 times smaller than
the best of the other applications, and the probability that a random
walker hits a state that is not due to an error is 10\% smaller than
the best of the other applications.
%
These values suggest that Gmail offers more ways to accomplish tasks
included in the scenario, that comparably more of these ways do not
involve extra steps, that they require fewer steps, and that it might
be more difficult for novice users to exploit the most efficient
ways. Further inspections show that some differences are due to the
slightly different interaction structures adopted for uploading
messages. If Gmail didn't exist yet, by using \migtool its designers
could obtain these answers well before developing prototypes and
performing usability studies.

@@

The literature on using state-transition network representations for
specifying or analysing the behavior of UIs is vast. We conducted a
systematic-style literature review, using Google Scholar and queries
with combinations of these phrases: ``user interface'', ``path
analysis'', ``user trace'', ``interaction trace'', ``navigation'',
``markov chain'', ``state transition'', ``statechart'', ``metric'',
``measure''. For each query we analyzed title and abstract of the
first 50 hits and appraised their relevance based on whether the paper
discusses approaches for measuring user interfaces in the context of
usability and on the basis of a model based on state transitions. This
resulted in 78 full-text papers that were later on re-analyzed against
the same criterion, leading to several of papers mentioned below.
%
Several of the examined papers deal with testing user interfaces, and
problems related to generating execution traces as a means to assess
coverage of a test
suite~\cite{ammann2008introduction,briand2005:test:uml}. They were
excluded from the review.

Usage of state-transition networks to model UI behavior in order to
draw usability conclusions dates back at least to~\cite{parnas69}. In
it, Parnas claimed that several kinds of usability problems would not
occur if the designer adopted a design framework where states and
their transitions are made explicit.

In many cases \emph{statecharts} are used (a generalization of finite
state automata, with an expressive language that includes hierarchic
levels of abstraction, concurrent regions, states and pseudostates, an
extended state notion based on an arbitrary computational model).
Harel, the inventor of statecharts, gives an interesting retrospective
view of how they were invented, back in early eighties, and why they
were appealing also to non experts~\cite{harel09}.  For the sake of
brevity, we refrain from describing them here in detail; the
interested reader is referred to the UML standard~\cite{uml241} or
some of the textbooks that deal with them, like~\cite{samek09}.
%
Horrocks showed how statecharts can be  used to model and
specify the dynamics of typical desktop UIs~\cite{horrocks99}. While
providing many interesting insights on how and why one should use
statecharts to do so, this nice work does not address how such a
specification could be \emph{automatically} processed.
%
This idea was later on greatly expanded by
Thimbleby~\cite{thimbleby07} , which discusses the use of statecharts
as meta-models for user interfaces; they are seen as crucial
representations that allow a designer to fully appreciate how devices
behave. The overall claim is that ``If you don't understand the logic
conveyed by a statechart model of a user interface, then you don't
understand the behavior of the user interface''.

WebML~\cite{ceri:00} is one of the most successful model-driven
approaches to web development (UIs and backend systems), with
industrial traction and a large number of publications. The language
is based on state transitions and is targeted to automatic generation
of data-intensive web applications.  %WebML encompasses different
%meta-models, independent from the technologies used to deliver the
%final application. % The WebML
% meta-model provides relatively few primitive units that can be
% combined to generate quite complex interactions; the modeling language
% is thus powerful and concise.  
Many other similar approaches involve or are based on task or activity
models~\cite{mori02,silva03:umli,koch02:uwe,gomez01,melia08}.
%
A recent OMG standard, called Interaction Flow Modeling Language
(IFML)~\cite{ifml13}, derives from WebML and focuses specifically on
user interaction. IFML is a language for specifying the structure of a
user interface and its behavior. It offers most of the abstractions
that are available in statecharts, mixed with the ability to specify
so-called ``components'' that are used to display and manipulate data
(to display details of a item, to display or select lists of items, to
input an item).

A rather different route for the problem of generating UIs is followed
in~\cite{heymann2007formal}. They assume that the UI to be generated
is used to supervise and to monitor an underlying machine (\eg,
autopilot of a plane) which is modeled as a statechart. After assuming
that the behavior of the UI can be modeled as a statechart, they
devised an algorithm that checks whether the two models are
compatible, and that refines the UI model so that its states and
transitions are minimized while still allowing a correct manipulation
of the underlying machine. Application of such a technique leads to
UIs that are correct by-construction.

In~\cite{conversy07} the Interactive Cooperative Objects formalism is
discussed, which is based on Petri nets, yet another state transition
formalism.  It is used to enrich the ARINC 661 specification of
Cockpit Display System used in interactive cockpits of airplanes, so
that the semantics of widgets can be expressed and the behavior of a
UI be formally analyzed.

%  that are now under provides precise information
% for communication protocol between application (called User
% Applications) and user interface components (called widgets) as well
% as precise information about the widgets themselves. However, in ARINC
% 661, no information is given about the behaviour of these widgets and
% about the behaviour of an application made up of a set of such
% widgets.  The Interactive Cooperative Objects (ICOs) formalism is a
% formal description technique dedicated to the specification of
% interactive systems [4, 11]. It uses concepts borrowed from the
% object-oriented approach to describe the structural or static aspects
% of systems, and uses high-level Petri nets [8] to describe their
% dynamic or behavioural aspects.

Finite state representations have been used also as a conceptual
framework to write the code of widgets so that events and event
handlers in the UI can more easily be conceived, developed and
verified (\eg, \cite{appert2008swingstates}).

% in java, FSA used as a conceptual framework to write the code of
% widgets so that events and event handlers in the ui can more easily be
% conceived, developed and verified.

% They say: StateCharts were used in the StateMaster User Interface
% Management System [40], and a variant of them specifically tailored to
% designing user interfaces was used in the more recent HsmTk toolkit
% [18]. StateCharts however are significantly more complicated and hard
% to learn than plain state machines, and our experience is that user
% interface designers and developers have difficulties exploiting their
% power. Other approaches include Petri Nets [41], which have also been
% used to specify user interfaces, for example in the PetShop system
% [42]. Here too, the learning curve is steep, making the adoption of
% such a model by developers difficult.

% With SwingStates, any number of state machines can run simulateneously. A state machine
% can be active, i.e., handling the events it receives, or inactive,
% i.e., ignoring events

% We adopt a similar apporach, but using UML state machine (which are
% more powerful than FSA) for conceiving, guiding development, analysis
% and verification of the behavior of the entire app.

% Nebeling says:\cite{nebeling2011metrics}

% Despite the fact that screen sizes and average screen resolutions have
% dramatically increased over the past few years, little attention has
% been paid to the design of web sites for large, high-resolution
% displays that are now becoming increasingly used both in enterprise
% and consumer spaces. We present a study of how the visual area of the
% browser window is currently utilised by news web sites at different
% widescreen resolutions. The analysis includes measurements of space
% taken up by the article content, embedded ads and the remaining
% components as they appear in the viewport of the web browser. The
% results show that the spatial distribution of page elements does not
% scale well with larger viewing sizes, which leads to an increasing
% amount of unused screen real estate and unnecessary scrolling. We
% derive a number of device-sensitive metrics to measure the quality of
% web page layout in different viewing contexts, which can guide the
% design of flexible layout templates that scale effectively on large
% screens.

% The purpose of ARINC 661 specification (ARINC 661, 2002) is to define
% interfaces to a Cockpit Display System (CDS) used in interactive
% cockpits that are now under

% application of a formal description technique to the various elements
% of ARINC 661 specification within an industrial project. This formal
% description technique called Interactive Cooperative Objects defines
% in a precise and non-ambiguous way all the elements of ARINC 661
% specification. The application of the formal description techniques is
% shown on an interactive application called MPIA (Multi Purpose
% Interactive Application). Within this application, we present how ICO
% are used for describing interactive widgets, User Applications and
% User Interface servers (in charge of interaction techniques). The
% emphasis is put on the model-based management of the feel of the
% applications allowing rapid prototyping of the external presentation
% and the interaction techniques.

% The Interactive Cooperative Objects (ICOs) formalism is a formal
% description technique dedicated to the specification of interactive
% systems [4, 11]. It uses concepts borrowed from the object-oriented
% approach to describe the structural or static aspects of systems, and
% uses high-level Petri nets [8] to describe their dynamic or
% behavioural aspects.

% The Interactive Cooperative Objects (ICOs) formalism is a formal
% description technique dedicated to the specification of interactive
% systems [4, 11]. It uses concepts borrowed from the object-oriented
% approach to describe the structural or static aspects of systems, and
% uses high-level Petri nets [8] to describe their dynamic or
% behavioural aspects.

The issue of measuring UI/interaction properties based on state
transition representations is not so well covered in the literature. A
review of usability measuring practices~\cite{hornbaek2006current},
under the headings \emph{measures of efficiency/usage patterns}, lists
only number of keystrokes, mouse clicks, and visited objects as
possible metrics. The notion of \emph{deviation from optimal solution}
is discussed only in the context of tools for route planning in 3D
navigation. In this paper we provide our own definition of deviation,
which applies to interaction with the UI; we provide also our notion
of execution traces across states of the UI, and the length of these
traces can be used as a measure of efficiency.

Lostness~\cite{smith1996lostness,otter2000lostness} is one of the few
metrics that were defined to measure the degree to which users become
lost in the information space, and therefore considers the notion of
\emph{deviation}. Defined specifically for hypertext systems, lostness
is a user performance measure that considers among others the number
of visited nodes, the number of different nodes that were visited
nodes, and the number of required nodes. This measure of efficiency is
usually applied to traces of actual users, and is argued to be
suitable for hypertext systems because the predominant task is
browsing information, rather than trying to achieve specific goals. It
therefore relaxes three assumptions: that there is a task to complete,
that there is a correct way to carry it out, that the purpose of the
system is to support users to carry out their tasks. The lostness
metric can be easily computed on top of the data that we use with our
approach; it would be yet another metric that can be provided to
assess quality of a UI. Notice that in our case we do not need to user
data to compute it.

A usability analysis method capable to analytically predict task
completion times from a storyboard of the UI is
CogTool~\cite{bellamy2011cogtool}. CogTool relies on the ACT-R
cognitive modeling engine, and allows a designer to setup a
low-fidelity prototype of the UI. After deciding which interaction
modality and which widgets are used to implement actions, the designer
gets a validated estimation of how long an experienced user would
spend on each step. CogTool takes care of adding extra ``mental''
steps, such as ``think'' steps, before certain patterns of provided
steps, according to the cognitive theory underlying ACT-R. As a
result, users of CogTool obtain the breakdown of the times required by
each of the steps. Our method is weaker in terms of cognitive validity
and in terms of precision of the output: it does not provide expected
completion times. However, with our method one can analyze a large
part of the UI, get information about possible problems in some areas,
and only then devolve more resources in building storyboards and
in making assumptions regarding widgets so that specific execution
paths previously identified can be analyzed with CogTool. In a sense,
the output of our method could be used to make informed decisions as
to what to analyze with CogTool.

Markov models, \ie\ directed graphs where edges leaving a vertex are
associated to a probability distribution, were used
in~\cite{thimbleby2001markov}, as a means to perform usability
analysis as early as possible, even before building prototypes of the
UI. Vertices represent states of the UI and edges correspond to user
actions (such as pressing a push-down button).  Probabilities can be
used as a model of the skills or knowledge of the user: equiprobable
actions correspond to a knowledge-free user, whereas when some actions
have a very low probability it means that for that user the action is
unlikely to be executed (perhaps because little known). Simple
mathematical operations on the transition matrix of the model give the
probability that after $n$ steps from a given initial state, the UI is
in a given state. With Markov models, by manipulating probabilities,
the designer can plot the number of required steps as a function of
how close the probabilities are to the designer's ``perfect''
knowledge.  Examples discussed in the paper cover several devices,
ranging from a simple torch (with 4 states), a microwave cooker (6
states), a mobile phone (152 states), but all being push-down devices
with a fixed set of buttons. This is obviously not the case for UIs of
information systems, where some buttons may or may not be ``painted''
on the screen. This makes it more difficult to specify the transition
matrix. Another limitation of this approach is that probabilities are
assumed to be constant, corresponding to users that are not learning
while interacting with the system.
%
Our approach is based on statecharts, a language that in practice is
much more concise than Markov models, making it easier to specify the
UI behavior, especially in cases where the set of buttons change over
time. While our examples do not make use of probabilities, this is
very easy to cope with (see the Discussion at the end of the paper for
some of the benefits that doing this could bring). Similarly
to~\cite{thimbleby2001markov}, our approach could be used when
conservative results that do not rely on psychological assumptions are
sought.

centrality measures

A discussion of social network analysis metrics applied to interaction
design is provided in~\cite{thimbleby09:sna:eics09}. Once more, a UI
is modeled in terms of directed graphs (vertices are states and edges
are actions), and various centrality metrics are used to draw
conclusions that bear upon usability. For example, centrality measures
(such as the Sabidussi,  Jordan, or betweenness) can be used to
identify states that are good places to start from to get to other
states. Other metrics, such as edge betweenness, can be used to
identify actions that are important because most of the shortest
paths go through these actions. The paper presents compelling examples
of using this technique to identify shortcomings in the design of
infusion pumps. 
%
In our work we automatically generate graphs from statechart models,
and tried to use these metrics. We were not able to draw sensible
conclusions from the values we obtained (for example from the models
presented below). One possible explanation may rest on the different
types of models: in our case they reflect the variety and flexibility
with which ``buttons'' can be used in modern web applications. For
infusion pumps the UI is more constrained in how a task can be carried
on.

% Voida\cite{voida} says:
% We describe an alternative model for organizing the
% desktop interface-activity-based computing-and
% identify a series of high-level system requirements for
% interfaces that use activity as their primary organizing
% principle.

% We could mention this work when we introduce scenarios, and say that
% people inventing these new metaphors could take advantage of our
% metrics if they model 2+ apps.

% Eg. they say: Requirement 2. Activity-based systems should provide
% lightweight mechanisms to create, change, and alter
% activities, since heavyweight interaction techniques are
% likely to deter adoption and use.

% eg Requirement 7. Because information sharing is a
% “common case” in knowledge work, lightweight sharing
% capabilities should be integrated directly as a first-class
% interaction technique.

@@

outline:
- modeling, statecharts
- flattening, ixgraph generation
- scenario def
- generation of traces
- analysis

modeling:
- statecharts details

flattening state machine \cite{briand2005:test:uml}
-

Notice that a scenario specifies only the desired occurrences of user
actions, not the necessary ones. For example, the model specifies that
in order to view a conversation while reading another one, one has to
\texttt{goBack} to the inbox first. It is the task of \migtool to
unfold cyles in the graph and compute all the  paths that connect the
desired user actions.

@@

An important issue underlying \migtool is the modeling effort that is
needed upfront. Our experience, based on several case studies and some
industrial examples, is that models do not need to be complete
representations of the behavior of the application under
study. By following an agile modeling approach~\cite{ambler02}, models can
be easily developed by one person in less than one day, using any UML
capable design tool. Even more complex models (in our experience up to
150 states and 350 transitions) can be developed and verified in 2-3
days by one person.
Experience in using statecharts to model behavior of UIs is needed
though. Useful suggestions are given in~\cite{horrocks99,thimbleby07}.

UML state diagrams provide a very expressive language, well suited to
specify behavior of UIs based on descrete events. Evidence of this can
also be found in recent OMG standards, such as
IFML~\cite{ifml13}. Even though there are fundamental limits
(inability to handle undo/redo's - because that goes beyond a finite
state representation; inability to handle customizable toolbars -
because that requires models that change at runtime; inability to
handle perceptive UIs - because they are not well suited to be
described in terms of descrete states), in many practical cases they
can be be isolated and/or left aside).

Expressivity of the modeling language means that different designers
are likely to produce different models for the same UI. As a
consequence, metrics computed by \migtool do depend on different
modeling choices. It is worth mentioning, though, that because of the
flattening process, several differences are reduced (for example,
those dealing with using a different hierarchy of states, or with
distributing differently concurrent regions across states), and such a
sensitivity is correspondingly reduced. At the moment, however, we do
not have hard evidence to back this claim.

Differently from other model-based approaches, such as those based on
IFML, \migtool uses \emph{only} a model of the behavior of the
UI. Designers do not need to cope with data modeling, nor with
decisions dealing with presentation. In a sense, \migtool uses only the
\emph{controller} part of the Model-View-Controller paradigm that is
adopted when developing UIs. Therefore, the designer using \migtool to
perform analysis is free from other concerns that in the end affect
usability, and the conclusions that are derived with \migtool can be
combined with other results \emph{after} the analysis is
performed. This also means that no effort needs to be directed on
developing data and presentation specifications/implementations.

In terms of validity of conclusions obtained through \migtool, we can
say that because they are devoid of user behavior assumptions (such as
preferences, skills, interpretations, ergonomic constraints) they are
very general. On the other hand, predictions based on \migtool metrics
are also generic because they do not consider data and presentation
aspects, nor user-related factors. For example, it is unfeasible to
use \migtool to predict the time needed by a user to complete a
scenario. However, \migtool can be used to analyze the whole
interaction design and gather data to inform more specific analyses
that could be performed, for example, with CogTool~\cite{john04cogtool}.

\migtool is part of a suite of model-based tools for designing,
analysing and testing user interfaces developed by the company
ANONYMIZED. Development of a web-based user interface of \migtool is
underway; the company plans to make it freely available for research
purposes.
